A Comparison of R Tools for Nonlinear Least Squares Modeling

### 2023-11-13 JNABresubmission20231101/ComparingNLStools.tex
resub 231101

% !TeX root = RJwrapper.tex
\title{A Comparison of R Tools for Nonlinear Least Squares Modeling}
\author{by John C. Nash and Arkajyoti Bhattacharjee}

@@

\subsection{\texorpdfstring{The \texttt{nls()} function}{The nls() function}}\label{the-nls-function}

@@

\subsection{Scope of our comparison}\label{scope-of-our-comparison}

The tools we will consider are the base-R \texttt{nls()} function
and some packages that are available on the CRAN repository.
Of these, we will pay particular attention to \textbf{nlsr} (John C Nash and Duncan Murdoch (2023)),
\textbf{minpack.lm} (Elzhov et al. (2012)), and \textbf{gslnls} (Chau (2023)) which are
general nonlinear least-squares solvers.

@@

\subsubsection{Principal messages}\label{principal-messages}

@@

For R \textbf{developers} we invite and encourage discussion of the design choices, since these
have downstream implications for ease of use, adaptation to new features and
efficiency of ongoing maintenance.

\subsection{Some other CRAN packages for nonlinear modeling}\label{some-other-cran-packages-for-nonlinear-modeling}

\texttt{onls} (Andrej-Nikolai Spiess (2022)) -- The usual optimization in estimating nonlinear models
is the vertical difference between the dependent variable and the functional
model. \texttt{onls} minimizes the sum of \textbf{orthogonal} residuals. The objective is
therefore different and involves nontrivial extra calculation. A vignette with
the package and the blog article
\url{https://www.r-bloggers.com/2015/01/introducing-orthogonal-nonlinear-least-squares-regression-in-r/}
give some description with illustrative graphs. \texttt{onls} appears to be limited to
problems with one independent and one dependent variable. The Wikipedia article
\url{https://en.wikipedia.org/wiki/Total_least_squares} presents an overview of some
ideas, with references to the literature. The approach needs a wider
discussion and tutorial examples to allow its merits to be judged than can be
included here.

@@

\texttt{NISTnls} ((ed.) (2012)) -- This package provides R code and data for a set of
(numerically ill-conditioned) nonlinear least squares problems from the U.S. National
Institute for Standards and Technology. These may not represent real-world situations.

@@

\texttt{nlsic} (Sokol (2022)) -- This solves nonlinear least squares problems with optional
equality and/or inequality constraints. It is clearly \textbf{not} about modeling, and
the input and output are quite different from class \texttt{nls} methods. However, there do
not appear to be other R packages with these capabilities.

@@

\texttt{nlsmsn} (Prates, Lachos, and Garay (2021)) -- Fit univariate non-linear scale mixture of skew-normal(NL-SMSN)
regression, with details in Garay, Lachos, and Abanto-Valle (2011). The problem
here is to minimize an objective that is modified from the traditional sum of squared
residuals.

@@

\texttt{nlstac} (Rodriguez-Arias et al. (2020)) -- A set of functions implementing the algorithm described in
Torvisco, Rodrıguez-Arias, and Sanchez (2018) for fitting separable nonlinear regression curves. The
special class of problem for which this package is intended is an important and
difficult one. No vignette is provided, unfortunately.

@@

\texttt{optimx} (John C. Nash and Varadhan (2011)) -- This provides
optimizers that can be applied to minimize a nonlinear function which could be
a nonlinear sum of
squares. NOT generally recommended if nonlinear
least squares programs can be easily used, but provides a check and alternative
solvers.

\subsection{An illustrative example}\label{an-illustrative-example}

The Hobbs weed infestation problem (John C. Nash (1979, 120)) is a growth curve modeling
task that seems straightforward but is quite nasty. Its very succinct statement
provides the ``short reproducible example'' much requested on R mailing lists.
This is a real problem from a field researcher. The data and graph follow.

@@

We thank one of the referees for expressing the viewpoint that a nonlinear regression
approach would use background knowledge and graphs to obtain an understanding of
the general magnitude of the parameters. Indeed, the ``visual fitting'' approach
(John C. Nash and Velleman (1996)) mentioned later accords with this viewpoint, as does use
of the \textbf{Logistic3T} variant of the model, as well as discussions in Ross (1990),
Seber and Wild (1989), D. M. Bates and Watts (1988), and Gallant (1987).

Our emphasis on the software resilience to starting parameters that are, in the
sense of statistical modeling, ``silly'' comes from over half a century of dealing
with users whose interests and understanding are very far from those of statistical
modelers. Thus we seek methods and codes that obtain reasonable answers under
highly unfavourable conditions. Nevertheless, a proper approach to
nonlinear modeling is to apply all available knowledge to the task.

@@

\[   Asym =  b_1 = 100 * c_1 \]
\[ exp(xmid/scal)  =  b_2 = 10 * c_2 \]
\[ 1/scal  =  b_3 = 0.1*c3 \]
To allow for simpler discussion, let us say that the parameters form a (named) vector
\(p\) and the model function is called \(model(p)\). The residuals can be written
either as \(r_1 = y - model(p)\) or \(r = model(p) - y\)
since their sum of squares has an identical value. The second form
avoids a potential sign error if we need to evaluate derivatives.

@@

Let us consider there are \(n\) parameters and \(m\) residuals.
The loss function is
\[ S(p) = r' r = \sum_{i=1}^m { r_i^2 }\]
The gradient of \(S(p)\) is
\[ g = 2 * J' r\]
where the Jacobian \(J\) is given by elements
\[ J_{i,j} = \partial r_i / \partial p_j \]
and the Hessian is defined by elements
\[ H_{i,j} = \partial ^2 S(p) / {\partial p_i \partial p_j} \]
If we expand the Hessian for nonlinear least squares problems, we find
\[  0.5 * H_{i,j} = \sum_{k = 1}^m{ J_{k,i} J_{k,j}}  + \sum_{k = 1}^m {r_k * \partial{r_k}/{\partial p_i \partial p_j}}\]
Let us use \(D_{i,j}\) for the elements of the second term of this expression.
What is generally called \textbf{Newton's method} for function minimization tries
to set the gradient to zero (to find a stationary point of the function \(S(p)\)).
This leads to \textbf{Newton's equation}
\[ H \delta = -g \]
Given a set of parameters \(p\), we solve this equation for \(\delta\), adjust \(p\)
to \(p + \delta\) and iterate, hopefully to converge on a solution.
Applying this to a sum of squares problem gives
\[ 0.5 * H \delta = (J'J + D) \delta = - J' r\]
In this expression, only the elements of \(D\) have second partial derivatives.
Gauss, attempting to model planetary orbits, had small residuals, and noted
that these multiplied the second partial derivatives of \(r\), so he approximated
\[ 0.5 * H \approx J' J\]
by assuming \(D \approx 0\). This results in the Gauss-Newton method where we
solve
\[ J' J \delta = - J' r\]
though we can avoid some loss of accuracy by NOT forming the inner product matrix
\(J' J\) and solving the linear least squares matrix problem
\[ J \delta \approx -r \]
by one of several matrix decomposition methods.

@@

\subsubsection{Problem setup}\label{problem-setup}

Let us specify in R the three model formulas and set some starting
values for parameters. These starting points are NOT equivalent and
are deliberately crude choices. Workers performing many calculations
of a similar nature should try to provide good starting points to reduce
computation time and avoid finding a false solution.

\begin{verbatim}
# model formulas
frmu <- weed ~ b1/(1+b2*exp(-b3*tt))
frms <- weed ~ 100*c1/(1+10*c2*exp(-0.1*c3*tt))
frmt <- weed ~ Asym /(1 + exp((xmid-tt)/scal))
#
# Starting parameter sets
stu1<-c(b1=1, b2=1, b3=1)
sts1<-c(c1=1, c2=1, c3=1)
stt1<-c(Asym=1, xmid=1, scal=1)
\end{verbatim}

@@

It is also possible to provide R functions for the residual and Jacobian.
This is usually much more work for the user if the formula setup is possible.
To illustrate, we show the functions for the unscaled 3 parameter logistic.
The particular form of these explicit residual and Jacobian functions comes from
their translation from BASIC codes of the 1970s, as adapted in John C. Nash and Walker-Smith (1987). The use
of the Laplace form of the residual and the inclusion of data within the functions
reflects choices of that era that are at odds with current practice. Some users
do still want or need to provide problems as explicit functions, particularly
for problems that are NOT regressions. For example, the Rosenbrock banana-valley
test problem can be provided this way, where the two ``residuals'' are different
functional forms.

\begin{verbatim}
# Logistic3U
hobbs.res  <-  function(x){ # unscaled Hobbs weeds problem -- residual
  if(length(x) != 3) stop("hobbs.res -- parameter vector n!=3")
  y  <-  c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192, 31.443, 
           38.558, 50.156, 62.948, 75.995, 91.972)
  tt  <-  1:12
  res  <-  x[1]/(1+x[2]*exp(-x[3]*tt)) - y
  # Note: this form of the residual, from Laplace (1788) in the form
  # of "fitted - observed" has been used in our software for half
  # a century, but sometimes offends users of the more traditional
  # "observed - fitted" form
}

hobbs.jac  <-  function(x) { # unscaled Hobbs weeds problem -- Jacobian
  jj  <-  matrix(0.0, 12, 3)
  tt  <-  1:12
  yy  <-  exp(-x[3]*tt)
  zz  <-  1.0/(1+x[2]*yy)
  jj[tt,1]   <-   zz
  jj[tt,2]   <-   -x[1]*zz*zz*yy
  jj[tt,3]   <-   x[1]*zz*zz*yy*x[2]*tt
  attr(jj, "gradient") <- jj
  jj
}
\end{verbatim}

\subsection{Estimation of models specified as formulas}\label{estimation-of-models-specified-as-formulas}

@@

\subsubsection{Solution attempts with nlsr}\label{solution-attempts-with-nlsr}

\begin{verbatim}
#> unlx1:
\end{verbatim}

\begin{verbatim}
#> residual sumsquares =  2.5873  on  12 observations
#>     after  19    Jacobian and  25 function evaluations
#>   name            coeff          SE       tstat      pval      gradient    JSingval   
#> b1               196.186         11.31      17.35  3.167e-08  -4.859e-09        1011  
#> b2               49.0916         1.688      29.08  3.284e-10  -3.099e-08      0.4605  
#> b3               0.31357      0.006863      45.69  5.768e-12   2.305e-06     0.04714
\end{verbatim}

@@

Though we have found solutions, the Jacobian is essentially singular as
shown by its singular
values. Note that these are \textbf{displayed} by package \texttt{nlsr} in a single column in
the output to provide a compact layout, but the values do \textbf{NOT} correspond to
the individual parameters in
whose row they appear; they are a property of the whole problem.

\subsubsection{Solution attempts with minpack.lm}\label{solution-attempts-with-minpack.lm}

@@

\subsubsection{Solution attempts with gslnls}\label{solution-attempts-with-gslnls}

@@

\subsubsection{Comparison notes for formula-setup solutions}\label{comparison-notes-for-formula-setup-solutions}

\texttt{nlsr::nlxb()} uses \texttt{print()} to output standard errors and
singular values of the Jacobian (for diagnostic purposes). By contrast,
\texttt{minpack.lm::nlsLM()} and \texttt{nls()}
use \texttt{summary()}, which does NOT display the sum of squares, while \texttt{print()}
gives the sum of squares, but not the standard error of the residuals.

@@

\begin{verbatim}
stspecial<- c(Asym = 35.532,  xmid = 43376,  scal = -2935.4)
badstart<-nlxb(formula=frmt, start=stspecial, data=weeddf)
print(badstart)
\end{verbatim}

@@

\subsection{Functional specification of problems}\label{functional-specification-of-problems}

We illustrate how to solve nonlinear least squares problems using
a function to define the residual. Note that \texttt{gsl\_nls()} requires
a vector \texttt{y} that is the expected value of what we have called the
residual, but in this case, is actually the model. \texttt{gsl\_nls} uses
a numerical approximation for the Jacobian if the argument \texttt{jac}
is missing. Note function \texttt{nlsr::pnlslm()} for a 1-line display of
the results of \texttt{minpack.lm::nls.lm()}.

@@

@@

@@

@@

\subsection{Design goals, termination tests, and output objects}\label{design-goals-termination-tests-and-output-objects}

@@

The design goals of the different tools may also be revealed in the so-called
``convergence tests'' for the iterative solvers. In the manual page for \texttt{nls()}
in R 4.0.0 there was the warning:

\begin{quote}
\emph{Do not use \texttt{nls} on artificial ``zero-residual'' data.}
\end{quote}

with suggested addition of small perturbations to the data. This admits
\texttt{nls()} could not solve well-posed problems unless data is polluted
with errors. Zero-residual problems are not always artificial,
since problems in function approximation and nonlinear equations can be
approached with nonlinear least squares. Fortunately, a small adjustment
to the ``termination test'' for the \textbf{program}, rather than for the
``convergence'' of the underlying \textbf{algorithm}, fixes the defect. The test
is the Relative Offset Convergence
Criterion (see Bates, Douglas M. and Watts, Donald G. (1981)). This scales an estimated reduction in the loss
function by its current value. If the loss function is very small, we are close
to a zero-divide. Adding a small quantity to the divisor avoids trouble.
In 2021, one of us (J. Nash) proposed that \texttt{nls.control()} have an
additional parameter \texttt{scaleOffset} with a default value of zero. Setting it to a
small number -- 1.0 is a reasonable choice -- allows small-residual problems
(i.e., near-exact fits) to be dealt with easily. We call this the
\textbf{safeguarded relative offset convergence criterion}, and it has been in \texttt{nlsr}
since it was introduced. The default value
gives the legacy behavior. This improvement has been in the R distributed code
since version 4.1.0.

@@

Termination after what may be considered excessive computation is also important.
\texttt{nls()} stops after \texttt{maxiter} ``iterations''. The meaning of ``iteration'' may require an
examination of the code. We terminate execution when the number of residual or
Jacobian evaluations exceed set limits. Generally, we prefer larger limits than the
default \texttt{maxiter\ =\ 50} of \texttt{nls()} to avoid stopping early.

\subsubsection{Returned results of nls() and other tools}\label{returned-results-of-nls-and-other-tools}

As mentioned, the output of \texttt{nls()}, \texttt{minpack.lm::nlsLM()}, or
\texttt{gslnls::gsl\_nls()} is an object of class ``nls'' which has a quite rich structure
described in the manual files or revealed by applying the \texttt{str()} function to the
result of \texttt{nls()}. The complexity of this object is a challenge to users. Let us
use for example \texttt{result\ \textless{}-\ snlm1} as the returned object from \texttt{nlsLM()} for the
Logistic3S problem. . The \texttt{data} return element
is an R symbol. To actually access the data from this element, we need to use
the syntax:

\begin{verbatim}
eval(parse(text=result$data))
\end{verbatim}

However, if the call is made with \texttt{model=TRUE}, then there is a returned element
\texttt{model} which contains the data, and we can list its contents using:

@@

\subsubsection{When to compute ancillary information}\label{when-to-compute-ancillary-information}

Tools that produce a class \texttt{nls} output object create a rich set of functions
and structures that are then used in a variety of modeling tasks, including the
least squares solution. By contrast,
\texttt{nlsr} computes quantities as they are requested or needed, with additional
features in separate functions. For example, the singular values of the Jacobian
are actually computed in the \texttt{print} and \texttt{summary} methods for the result.
These two approaches
lead to different consequences for performance and how features are
provided. \texttt{nlsr} has antecedents in the methods of John C. Nash (1979), where storage for
data and programs was at a ridiculous premium in the small computers of the era.
Thus the code in \texttt{nlsr} is likely of value for workers to copy and modify
for customized tools.

\subsection{Jacobian calculation}\label{jacobian-calculation}

@@

Using the ``gradient'' attribute of the output of the Jacobian function to hold
the Jacobian matrix lets us embed this in the \textbf{residual} function as well,
so that the call to \texttt{nlsr::nlfb()} can be made with the same name used for
both residual and Jacobian function arguments. This programming trick saves a
lot of trouble for the package developer, but it can be a nuisance for users
trying to understand the code.

As far as we can understand the logic in \texttt{nls()}, the Jacobian computation during
parameter estimation is carried out within the called C-language program
and its wrapper R code function \texttt{numericDeriv()}, part of
\texttt{./src/library/stats/R/nls.R}
in the R distribution source code. This is used to provide Jacobian information in
the \texttt{nlsModel()} and \texttt{nlsModel.plinear()} functions, which are \textbf{not} exported for
general use. \texttt{gsl\_nls()} also appears to use \texttt{numericDeriv()}.

@@

\begin{itemize}
\item
  A pasture regrowth problem (Huet et al. (2004), page 1, based on Ratkowsky (1983)) has a
  poorly conditioned Jacobian and \texttt{nls()} fails with ``singular gradient''.
  Worse, numerical approximation to the Jacobian gives the error
  ``singular gradient matrix at initial parameter estimates'' for \texttt{minpack.lm::nlsLM}
  so that the Marquardt stabilization is unable to take effect, while the analytic
  derivatives of \texttt{nlsr::nlxb} give a solution.
\item
  Karl Schilling (private communication) provided an example where a model specified
  with the formula \texttt{y\ \textasciitilde{}\ a\ *\ (x\ ˆ\ b)} causes \texttt{nlsr::nlxb} to fail because the partial
  derivative w.r.t. \texttt{b} is \texttt{a\ *\ (x\^{}b\ *\ log(x))}. If there is data for which \texttt{x\ =\ 0},
  the derivative is undefined, but the model can be computed. In such cases,
  we observed that \texttt{nls()} and
  \texttt{minpack.lm::nlsLM} found a solution, though this seems to be a lucky accident.
\end{itemize}

>>>>

% !TeX root = RJwrapper.tex
\title{A Comparison of R Tools for Nonlinear Least Squares Modeling}

\author{by John C. Nash and Arkajyoti Bhattacharjee}

@@

\hypertarget{the-nls-function}{%
\section{\texorpdfstring{The \texttt{nls()} function}{The nls() function}}\label{the-nls-function}}

@@

\hypertarget{scope-of-our-comparison}{%
\section{Scope of our comparison}\label{scope-of-our-comparison}}

The tools we will consider are the base-R \texttt{nls()} function
and some packages that are available on the CRAN repository.
Of these, we will pay particular attention to \textbf{nlsr} (John C. Nash and Murdoch (2023)),
\textbf{minpack.lm} (Elzhov et al. (2012)), and \textbf{gslnls} (Chau (2023)) which are
general nonlinear least-squares solvers.

@@

\hypertarget{principal-messages}{%
\subsection{Principal messages}\label{principal-messages}}

@@

For R \textbf{developers}, we invite and encourage discussion of the design choices, since these
have downstream implications for ease of use, adaptation to new features, and
efficiency of ongoing maintenance.

\hypertarget{some-other-cran-packages-for-nonlinear-modeling}{%
\section{Some other CRAN packages for nonlinear modeling}\label{some-other-cran-packages-for-nonlinear-modeling}}

\texttt{onls} (Spiess, Andrej-Nikolai (2022)) -- The usual optimization in estimating nonlinear models
is the vertical difference between the dependent variable and the functional
model. \texttt{onls} minimizes the sum of \textbf{orthogonal} residuals. The objective is
therefore different and involves nontrivial extra calculation. A vignette with
the package and the blog article
\url{https://www.r-bloggers.com/2015/01/introducing-orthogonal-nonlinear-least-squares-regression-in-r/}
give some description with illustrative graphs. \texttt{onls} appears to be limited to
problems with one independent and one dependent variable. The Wikipedia article
\url{https://en.wikipedia.org/wiki/Total_least_squares} presents an overview of some
ideas, with references to the literature. The approach needs a wider
discussion and tutorial examples to allow its merits to be judged than can be
included here.

@@

\texttt{NISTnls} (D. M. Bates (2012)) -- This package provides R code and data for a set of
(numerically ill-conditioned) nonlinear least squares problems from the U.S. National
Institute for Standards and Technology. These may not represent real-world situations.

@@

\texttt{nlsic} (Sokol (2022)) -- This solves nonlinear least squares problems with optional
equality and/or inequality constraints. It is clearly not about modeling, and
the input and output are quite different from class \texttt{nls} methods. However, there do
not appear to be other R packages with these capabilities.

@@

\texttt{nlsmsn} (Prates, Lachos, and Garay (2021)) -- Fit univariate non-linear scale mixture of skew-normal (NL-SMSN)
regression, with details in Garay, Lachos, and Abanto-Valle (2011). The problem
here is to minimize an objective that is modified from the traditional sum of squared
residuals.

@@

\texttt{nlstac} (Rodriguez-Arias et al. (2020)) -- A set of functions implementing the algorithm described in
Torvisco, Rodriguez-Arias, and Sanchez (2018) for fitting separable nonlinear regression curves. The
special class of problem for which this package is intended is an important and
difficult one. No vignette is provided, unfortunately.

@@

\texttt{optimx} (John C. Nash and Varadhan (2011)) -- This provides
optimizers that can be applied to minimize a nonlinear function which could be
a nonlinear sum of squares. Not generally recommended if nonlinear
least squares programs can be easily used, but provides a check and alternative
solvers.

\hypertarget{an-illustrative-example}{%
\section{An illustrative example}\label{an-illustrative-example}}

The Hobbs weed infestation problem (John C. Nash (1979, 120)) is a growth-curve modeling
task that seems straightforward but is quite nasty. Its very succinct statement
provides the ``short reproducible example'' much requested on R mailing lists.
This is a real problem from a field researcher. The data and graph follow.

@@

While model estimation is increasingly automated, nonlinear regression
should use background knowledge and graphs to obtain an understanding of
the general magnitude of the parameters. Indeed, the ``visual fitting'' approach
(John C. Nash and Velleman (1996)) mentioned later accords with this viewpoint, as does use
of the \textbf{Logistic3T} variant of the model, as well as discussions in Ross (1990),
Seber and Wild (1989), D. M. Bates and Watts (1988), and Gallant (1987).
Our emphasis on the software resilience to starting parameters that are, in the
sense of statistical modeling, ``silly'' comes from over half a century of dealing
with users whose interests and understanding are very far from those of statistical
modelers. Thus, we seek methods and codes that obtain reasonable answers under
highly unfavourable conditions. Nevertheless, a proper approach to
nonlinear modeling is to apply all available knowledge to the task.

@@

\[   Asym =  b_1 = 100 * c_1 \]
\[ exp(xmid/scal)  =  b_2 = 10 * c_2 \]
\[ 1/scal  =  b_3 = 0.1*c3 \]
To allow for simpler discussion, let us say that the parameters form a (named) vector
\(p\) and the model function is called \(model(p)\) with \(r = y - model(p)\).

@@

Let us consider there are \(n\) parameters and \(m\) residuals.
The loss function is
\[ S(p) = r' r = \sum_{i=1}^m { r_i^2 }\]
The gradient of \(S(p)\) is
\[ g = 2 * J' r\]
where the Jacobian \(J\) is given by elements
\[ J_{i,j} = \partial r_i / \partial p_j \]
and the Hessian is defined by elements
\[ H_{i,j} = \partial ^2 S(p) / {\partial p_i \partial p_j} \]
If we expand the Hessian for nonlinear least squares problems, we find
\[  0.5 * H_{i,j} = \sum_{k = 1}^m{ J_{k,i} J_{k,j}}  + \sum_{k = 1}^m {r_k * \partial{r_k}/{\partial p_i \partial p_j}}\]
Let us use \(D_{i,j}\) for the elements of the second term of this expression.
What is generally called \textbf{Newton's method} for function minimization tries
to set the gradient to zero (to find a stationary point of the function \(S(p)\)).
This leads to \textbf{Newton's equation}
\[ H \delta = -g \]
Given a set of parameters \(p\), we solve this equation for \(\delta\), adjust \(p\)
to \(p + \delta\) and iterate, hopefully to converge on a solution.
Applying this to a sum of squares problem gives
\[ 0.5 * H \delta = (J'J + D) \delta = - J' r\]
In this expression, only the elements of \(D\) have second partial derivatives.
Gauss, attempting to model planetary orbits, had small residuals, and noted
that these multiplied the second partial derivatives of \(r\), so he approximated
\[ 0.5 * H \approx J' J\]
by assuming \(D \approx 0\). This results in the Gauss-Newton method where we
solve
\[ J' J \delta = - J' r\]
though we can avoid some loss of accuracy by not forming the inner product matrix
\(J' J\) and solving the linear least squares matrix problem
\[ J \delta \approx -r \]
by one of several matrix decomposition methods.

@@

\hypertarget{problem-setup}{%
\subsection{Problem setup}\label{problem-setup}}

Let us specify in R the three model formulas and set some starting
values for parameters. These starting points are not equivalent and
are deliberately crude choices. Workers performing many calculations
of a similar nature should try to provide good starting points to reduce
computation time and avoid finding a false solution.

\begin{verbatim}
# model formulas
frmu <- weed ~ b1 / (1 + b2 * exp(-b3 * tt))
frms <- weed ~ 100 * c1/(1 + 10* c2* exp(-0.1 * c3* tt))
frmt <- weed ~ Asym / (1 + exp((xmid - tt) / scal))
#
# Starting parameter sets
stu1 <- c(b1 = 1, b2 = 1, b3 = 1)
sts1 <- c(c1 = 1, c2 = 1, c3 = 1)
stt1 <- c(Asym = 1, xmid = 1, scal = 1)
\end{verbatim}

@@

It is also possible to provide R functions for the residual and Jacobian.
This is usually much more work for the user if the formula setup is possible.
To illustrate, we show the functions for the unscaled 3 parameter logistic.
The particular form of these explicit residual and Jacobian functions comes from
their translation from BASIC codes of the 1970s, as adapted in John C. Nash and Walker-Smith (1987). The use
of the Laplace form of the residual and the inclusion of data within the functions
reflects choices of that era that are at odds with current practice. Some users
do still want or need to provide problems as explicit functions, particularly
for problems that are not regressions. For example, the Rosenbrock banana-valley
test problem can be provided this way, where the two ``residuals'' are different
functional forms.

\begin{verbatim}
# Logistic3U
hobbs.res <- function(x){ # unscaled Hobbs weeds problem -- residual
  if(length(x) != 3) stop("hobbs.res -- parameter vector n!=3")
  y <- c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192, 31.443, 
           38.558, 50.156, 62.948, 75.995, 91.972)
  tt <- 1:12
  res <- x[1] / (1 + x[2] * exp(-x[3] * tt)) - y
  # Note: this form of the residual, from Laplace (1788) in the form
  # of "fitted - observed" has been used in our software for half
  # a century, but sometimes concerns users of the more traditional
  # "observed - fitted" form
}

hobbs.jac <- function(x) { # unscaled Hobbs weeds problem -- Jacobian
  jj <- matrix(0.0, 12, 3)
  tt <- 1:12
  yy <- exp(-x[3] * tt)
  zz <- 1.0 / (1 + x[2] * yy)
  jj[tt, 1] <- zz
  jj[tt, 2] <- -x[1] * zz * zz * yy
  jj[tt, 3] <- x[1] * zz * zz * yy * x[2] * tt
  attr(jj, "gradient") <- jj
  jj
}
\end{verbatim}

\hypertarget{estimation-of-models-specified-as-formulas}{%
\section{Estimation of models specified as formulas}\label{estimation-of-models-specified-as-formulas}}

@@

\hypertarget{solution-attempts-with-nlsr}{%
\subsection{Solution attempts with nlsr}\label{solution-attempts-with-nlsr}}

\begin{verbatim}
#> unlx1:residual sumsquares =  2.5873  on  12 observations
#>     after  19    Jacobian and  25 function evaluations
#>   name            coeff          SE       tstat      pval      gradient    JSingval   
#> b1               196.186         11.31      17.35  3.167e-08  -4.859e-09        1011  
#> b2               49.0916         1.688      29.08  3.284e-10  -3.099e-08      0.4605  
#> b3               0.31357      0.006863      45.69  5.768e-12   2.305e-06     0.04714
\end{verbatim}

@@

Though we have found solutions, the Jacobian is essentially singular as
shown by its singular
values. Note that these are displayed by package \texttt{nlsr} in a single column in
the output to provide a compact layout, but the values do not correspond to
the individual parameters in
whose row they appear; they are a property of the whole problem.

\hypertarget{solution-attempts-with-minpack.lm}{%
\subsection{Solution attempts with minpack.lm}\label{solution-attempts-with-minpack.lm}}

@@

\hypertarget{solution-attempts-with-gslnls}{%
\subsection{Solution attempts with gslnls}\label{solution-attempts-with-gslnls}}

@@

\hypertarget{comparison-notes-for-formula-setup-solutions}{%
\subsection{Comparison notes for formula-setup solutions}\label{comparison-notes-for-formula-setup-solutions}}

\texttt{nlsr::nlxb()} uses \texttt{print()} to output standard errors and
singular values of the Jacobian (for diagnostic purposes). By contrast,
\texttt{minpack.lm::nlsLM()} and \texttt{nls()}
use \texttt{summary()}, which does not display the sum of squares, while \texttt{print()}
gives the sum of squares, but not the standard error of the residuals.

@@

\begin{verbatim}
stspecial <- c(Asym = 35.532,  xmid = 43376,  scal = -2935.4)
badstart <- try(nlxb(formula = frmt, start = stspecial, data = weeddf))
if (! inherits(badstart, "try-error")) print(badstart)
\end{verbatim}

@@

\hypertarget{functional-specification-of-problems}{%
\section{Functional specification of problems}\label{functional-specification-of-problems}}

We illustrate how to solve nonlinear least squares problems using
a function to define the residual. Note that \texttt{gsl\_nls()} requires
a vector \texttt{y} that is the length of the vector returned by the
function supplied, e.g., \texttt{hobbs.res()}. \texttt{gsl\_nls} uses
a numerical approximation for the Jacobian if the argument \texttt{jac}
is missing. Note function \texttt{nlsr::pnlslm()} for a 1-line display of
the results of \texttt{minpack.lm::nls.lm()}.

\begin{verbatim}
#> hobnlfb<-nlfb(start=stu1, resfn=hobbs.res, jacfn=hobbs.jac)
\end{verbatim}

@@

\begin{verbatim}
#> hobnlm<-nls.lm(par=stu1, fn=hobbs.res, jac=hobbs.jac)
\end{verbatim}

@@

\begin{verbatim}
#> hobgsln<-gsl_nls(start=stu1, fn=hobbs.res, y=rep(0,12))
\end{verbatim}

@@

\begin{verbatim}
#> hobgsl<-gsl_nls(start=stu1, fn=hobbs.res, y=rep(0,12), jac=hobbs.jac)
\end{verbatim}

@@

\hypertarget{design-goals-termination-tests-and-output-objects}{%
\section{Design goals, termination tests, and output objects}\label{design-goals-termination-tests-and-output-objects}}

@@

The design goals of the different tools may also be revealed in the so-called
``convergence tests'' for the iterative solvers. In the manual page for \texttt{nls()}
in R 4.0.0, there was the warning:

\begin{quote}
\emph{Do not use \texttt{nls} on artificial ``zero-residual'' data}
\end{quote}

with suggested addition of small perturbations to the data. This admits
\texttt{nls()} could not solve well-posed problems unless data is polluted
with errors. Zero-residual problems are not always artificial,
since problems in function approximation and nonlinear equations can be
approached with nonlinear least squares. Fortunately, a small adjustment
to the ``termination test'' for the program, rather than for the
``convergence'' of the underlying algorithm, fixes the defect. The test
is the Relative Offset Convergence
Criterion (see D. M. Bates and Watts (1981)). This scales an estimated reduction in the loss
function by its current value. If the loss function is very small, we are close
to a zero-divide. Adding a small quantity to the divisor avoids trouble.
In 2021, one of us (J. Nash) proposed that \texttt{nls.control()} have an
additional parameter \texttt{scaleOffset} with a default value of zero. Setting it to a
small number -- 1.0 is a reasonable choice -- allows small-residual problems
(i.e., near-exact fits) to be dealt with easily. We call this the
\textbf{safeguarded relative offset convergence criterion}, and it has been in \texttt{nlsr}
since it was introduced. The default value
gives the legacy behavior. This improvement has been in the R distributed code
since version 4.1.0.

@@

Termination after what may be considered excessive computation is also important.
\texttt{nls()} stops after \texttt{maxiter} ``iterations''. The meaning of ``iteration'' may require an
examination of the code for the different algorithms. \texttt{nlsr} terminates execution
when the number of residual or Jacobian evaluations exceed set limits. Generally,
we prefer larger limits than the default \texttt{maxiter\ =\ 50} of \texttt{nls()} to avoid stopping
early, though this may result in some unnecessary computations.

\hypertarget{returned-results-of-nls-and-other-tools}{%
\subsection{Returned results of nls() and other tools}\label{returned-results-of-nls-and-other-tools}}

As mentioned, the output of \texttt{nls()}, \texttt{minpack.lm::nlsLM()}, or
\texttt{gslnls::gsl\_nls()} is an object of class ``nls'' which has a quite rich structure
described in the manual files or revealed by applying the \texttt{str()} function to the
result of \texttt{nls()}. The complexity of this object is a challenge to users. Let us
use for example \texttt{result\ \textless{}-\ snlm1} as the returned object from \texttt{nlsLM()} for the
Logistic3S problem. The \texttt{data} return element
is an R symbol. To actually access the data from this element, we need to use
the syntax:

\begin{verbatim}
eval(parse(text = result$data))
\end{verbatim}

However, if the call is made with \texttt{model\ =\ TRUE}, then there is a returned element
\texttt{model} which contains the data, and we can list its contents using:

@@

\hypertarget{when-to-compute-ancillary-information}{%
\subsection{When to compute ancillary information}\label{when-to-compute-ancillary-information}}

Tools that produce a class \texttt{nls} output object create a rich set of functions
and structures that are then used in a variety of modeling tasks, including the
least squares solution. By contrast,
\texttt{nlsr} computes quantities as they are requested or needed, with additional
features in separate functions. For example, the singular values of the Jacobian
are actually computed in the \texttt{print} and \texttt{summary} methods for the result.
These two approaches
lead to different consequences for performance and how features are
provided. \texttt{nlsr} has antecedents in the methods of John C. Nash (1979), where storage for
data and programs was at a ridiculous premium in the small computers of the era.
Thus, the code in \texttt{nlsr} is likely of value for workers to copy and modify
for customized tools.

\hypertarget{jacobian-calculation}{%
\section{Jacobian calculation}\label{jacobian-calculation}}

@@

Using the ``gradient'' attribute of the output of the Jacobian function to hold
the Jacobian matrix lets us embed this in the residual function as well,
so that the call to \texttt{nlsr::nlfb()} can be made with the same name used for
both residual and Jacobian function arguments. This programming trick saves a
lot of trouble for the package developer, but it can be a nuisance for users
trying to understand the code.

As far as we can understand the logic in \texttt{nls()}, the Jacobian computation during
parameter estimation is carried out within the called C-language program
and its wrapper R code function \texttt{numericDeriv()}, part of
\texttt{./src/library/stats/R/nls.R}
in the R distribution source code. This is used to provide Jacobian information in
the \texttt{nlsModel()} and \texttt{nlsModel.plinear()} functions, which are not exported for
general use. \texttt{gsl\_nls()} also appears to use \texttt{numericDeriv()}.

@@

\begin{itemize}
\item
  A pasture regrowth problem (Huet et al. (2004), page 1, based on Ratkowsky (1983)) has a
  poorly conditioned Jacobian and \texttt{nls()} fails with ``singular gradient''.
  Worse, numerical approximation to the Jacobian gives the error
  ``singular gradient matrix at initial parameter estimates'' for \texttt{minpack.lm::nlsLM}
  so that the Marquardt stabilization is unable to take effect, while the analytic
  derivatives of \texttt{nlsr::nlxb} give a solution.
\item
  Karl Schilling (private communication) provided an example where a model specified
  with the formula \texttt{y\ \textasciitilde{}\ a\ *\ (x\ ˆ\ b)} causes \texttt{nlsr::nlxb} to fail because the partial
  derivative w.r.t. \texttt{b} is \texttt{a\ *\ (x\ \^{}\ b\ *\ log(x))}. If there is data for which \texttt{x\ =\ 0},
  the derivative is undefined, but the model can be computed. In such cases,
  we observed that \texttt{nls()} and
  \texttt{minpack.lm::nlsLM} found a solution, though this seems to be a lucky accident.
\end{itemize}

\hypertarget{jacobian-code-in-selfstart-models}{%
\subsection{Jacobian code in selfStart models}\label{jacobian-code-in-selfstart-models}}
